# Reproducible Research Week 2 Assignment
=========================================

## Task 1: Loading and preprocessing the data
- Downloading the data
- Unzipping the original data file
- Loading the data into an 'R' object via the read.csv() funtion.
```{r load, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",destfile = "Rep Research Week2.zip")
data <- read.csv(unzip("Rep Research Week2.zip"))
```

## Task 2: What is mean total number of steps taken per day
2.1 Calculate the total number of steps taken per day
```{r totalsteps, include=TRUE}
data_noNA <- data[!is.na(data$'steps'),]

#using an SQL style query to create a dataframe containing average steps per day by date.
library(sqldf)
total_steps_perday <- sqldf("select sum(steps),date from data_noNA group by date")
colnames(total_steps_perday) <- c("steps","date")
#The total number of steps per day is:
total_steps_perday
```


2.2 Make a histogram of the total number of steps taken each day
```{r histogram}
library(ggplot2)
ggplot(data=total_steps_perday) + geom_histogram(mapping = aes(x= steps))
```

2.3 Calculate and report the mean and median of the total number of steps taken per day:
```{r mean}
mean <- mean(total_steps_perday$steps)
median <- median(total_steps_perday$steps)
```
The mean is `r mean`
The median is `r median`

## Task 3: What is the average daily activity pattern?
3.1 Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r timeseries}
average_steps_perinterval <- sqldf("select avg(steps),interval,count(*) from data_noNA group by interval")
colnames(average_steps_perinterval) <- c("avg_steps","interval","count")

## The followiing graph depicts thew timeseries plot of average no. of steps across various 5 minute intervals.
ggplot(average_steps_perinterval, aes(interval, avg_steps)) + geom_line() + ylab("Avg. no. of Steps") + xlab("interval")
```

3.2 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r max_average}
max_average_interval <- average_steps_perinterval[max(average_steps_perinterval$'avg_steps'),]
max_average_interval
```
The average max average of steps is `r max_average_interval$avg_steps`

## Task 4: Imputing missing values
4.1 Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r missing_values}
data_NAs <- data[is.na(data$'steps'),]
number_of_NAs <- nrow(data_NAs)
number_of_NAs
```
The total number of NAs is `r number_of_NAs`

4.2 Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r missing NAs}
replace_NAs <- data_NAs
average_steps_perday1 <- sqldf("select avg(steps),interval from data_noNA group by interval")
colnames(average_steps_perday1) <- c("steps2","interval")

replace_NAs_final <- sqldf("select a.steps, a.date, a.interval, b.steps2 from replace_NAs a left join average_steps_perday1 b on a.interval = b.interval")
replace_NAs_final1 <- as.data.frame(cbind(replace_NAs_final$steps2,as.character(replace_NAs_final$date),replace_NAs_final$interval))
colnames(replace_NAs_final1) <- colnames(data)
```
4.3 Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r reconstructed}
reconstructed <- rbind(data_noNA,replace_NAs_final1)
```

4.4 Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r Histogram_imputed}
library(ggplot2)
library(sqldf)
total_steps_imputed <- sqldf("select sum(steps),date from reconstructed group by date")
# The total number of steps per day (after imputing NAs) is:
total_steps_imputed
colnames(total_steps_imputed) <- c("steps","date")
# Histogram of total steps after imputing:
ggplot(data=total_steps_imputed) + geom_histogram(mapping = aes(x= steps))
```

## 5. Are there differences in activity patterns between weekdays and weekends?
5.1 Create a new factor variable in the dataset with two levels weekdays and weekends indicating whether a given date is a weekday or weekend day.
```{r weekdays}
weekdays <- reconstructed
weekdays[,"weekday"] <- NA #adding a blank column to use as a weekend/weekday indicator.

#using a custom function along with function sapply() to add a flag called "weekday" that indicates whether
#a row corresponds to a Weekend observation or a Weekday observation.
return_weekday <- function(x){   
  if(weekdays(x) %in% c("Saturday","Sunday"))
     {
       return("Weekend")
  }
     return("Weekday")
}

weekdays$weekday <- sapply(as.Date(weekdays$date), function(t) return_weekday(t))
weekdays$steps <- as.numeric(weekdays$steps) #updating the weekday derived, into the data frame
```
5.2 Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r timeseries2}
library(ggplot2)
library(sqldf)
plot1 <- sqldf("select avg(steps),interval, weekday from weekdays group by interval, weekday")
str(plot1)
colnames(plot1) <- c("steps","interval","weekday")
ggplot(plot1, aes(as.numeric(interval), steps)) + geom_line() + ylab("Avg. no. of Steps") + xlab("interval") + facet_wrap(~weekday, nrow = 2)
```